---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r, echo = TRUE}

```

## Loading and preprocessing the data
First, i loaded the data from activity.csv file into R.
```{r, echo = TRUE}
dataset <- read.csv(file="/home/kasia/RepData_PeerAssessment1/activity.csv", header=TRUE, sep=",")
```
In order to process or transform the data into a format suitable for my analysis, i took a quick look
at the data and its structure.
```{r, echo = TRUE}
head(dataset)
str(dataset)
```
I transformed 'date' column from factor type into date type.
```{r, echo = TRUE}
dataset$date <- as.Date(dataset$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
Since for this part of the assignment missing values in the dataset can be ignored, i extracted a subset containing only complete cases.
```{r, echo = TRUE}
complete_cases_only <- dataset[complete.cases(dataset),]
```
In order to prepare a histogram of the total number of steps taken each day, i used dplyr package to prepare the dataset. 
```{r, echo = TRUE, fig.width = 6, fig.height = 6, fig.path='figure/'}
library(dplyr)
total_steps_per_day_1 <- summarise( group_by(complete_cases_only, date), total_steps = sum(steps))

hist(total_steps_per_day_1$total_steps
      , xlab = "Total steps per day"
      , ylab = "Entries"
      , main = NULL
      , breaks = 10
      )

```

I calculated the mean and median of total number of steps taken per day.
```{r, echo = TRUE}
mean(total_steps_per_day_1$total_steps )
median(total_steps_per_day_1$total_steps )
```

## What is the average daily activity pattern?
To make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days, i prepared a dataset using dplyr package...
```{r, echo = TRUE}
avg_steps_per_interval <- summarise( group_by(dataset, interval), avg_steps = mean(steps, na.rm = TRUE))
```
... and used ggplot package to create a plot.
```{r, echo = TRUE, fig.width = 6, fig.height = 6, fig.path='figure/'}
library(ggplot2)

ggplot(avg_steps_per_interval,
        aes(x=interval,
            y=avg_steps)
  ) +
  geom_line() +  
  xlab("Interval") +
  ylab("Average steps")
```

I checked which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.
```{r, echo = TRUE}
avg_steps_per_interval[which.max(avg_steps_per_interval$avg_steps),]
```

## Imputing missing values
I calculated the total number of missing values in the dataset.
```{r, echo = TRUE}
nrow(dataset[!complete.cases(dataset),])
```
In order to fill in all of the missing values in the dataset, i used the mean for that 5-minute interval, calculated in the previous step. 

I created a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo = TRUE}
missing_values_filled <- dataset
for (i in 0:max(missing_values_filled$interval)){
    missing_values_filled[(!complete.cases(missing_values_filled) & missing_values_filled$interval == i) , 1 ]  <-  avg_steps_per_interval[avg_steps_per_interval$interval == i, 2]
}
```

For the dataset with missing values filled in, i made a histogram of the total number of steps taken each day.
```{r, echo = TRUE, fig.width = 6, fig.height = 6, fig.path='figure/'}
total_steps_per_day_2 <- summarise( group_by(missing_values_filled, date), total_steps = sum(steps))

hist(total_steps_per_day_2$total_steps
     , xlab = "Total steps per day"
     , ylab = "Entries"
     , main = NULL
     , breaks = 10
     )
```


The median total number of steps taken per day differ from the estimate from the first part of the assignment.  
```{r, echo = TRUE}
median(total_steps_per_day_2$total_steps )
```
The mean value does not differ from the estimate.
```{r, echo = TRUE}
mean(total_steps_per_day_2$total_steps )
```

In this dataset, mean and median values of total number of steps taken per day are equal (which was not the case in the dataset with missing values ignored).

## Are there differences in activity patterns between weekdays and weekends?
I created a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r, echo = TRUE}
missing_values_filled$day_of_week <- weekdays(missing_values_filled$date) 

missing_values_filled$day_type[ missing_values_filled$day_of_week == "sobota"] <- "weekend"
missing_values_filled$day_type[ missing_values_filled$day_of_week == "niedziela"] <- "weekend"
missing_values_filled$day_type[ ( missing_values_filled$day_of_week != "niedziela" & missing_values_filled$day_of_week != "sobota") ] <- "weekday"

missing_values_filled$day_type <- factor(missing_values_filled$day_type)
```

Finally i made a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.
```{r, echo = TRUE, fig.width = 6, fig.height = 6, fig.path='figure/'}
avg_steps_per_interval_by_day_type <- summarise( group_by(missing_values_filled, interval, day_type), avg_steps = mean(steps))

library(ggplot2)

ggplot(avg_steps_per_interval_by_day_type,
       aes(x=interval,
           y=avg_steps)
  ) +
  geom_line() +  
  xlab("Interval") +
  ylab("Average steps") +
  facet_grid(day_type ~ .)
```

